### 2021-07-07 ack.tex
this is a nightmare

Most of all, to mom Calisa Pierce, dad Jim Pierce, my sisters Kelsey and Kim, and my grandmother Wilma Steele, you've formed all of the best parts of my personality and continually guided me toward this point, and it's impossible to explain how thankful I am for that. Regarding family I will also always be thankful for my great aunt Alice who graciously paid for my undergraduate education.

@@

Thanks also to my professors at Southern WV Community College (SWVCC) and West Virginia University (WVU) who fostered my way of thinking and provided many of the tools I used for the research in this thesis.
At SWVCC, I'd like to thank especially Mindy Saunders, Tex Wood, Anne Klein, Charles Wood, Don Saunders, and George Trimble. Mindy built my math and physics foundation with her incredibly effective teaching, Anne showed me there's invisible order in everything, and Tex demonstrated that art will always lead science. Everyone encouraged me to continue my education. SWVCC was a personal transformation and a great privilege.

@@

To my extended cohort at WVU physics, Payne, Megat, Collins, Scott, Dustin, Rice, Wolfe, Samet, Craig, April, Evan, Larry, Robert, Stephen, Rick, Matt, Caleb, John, and all of you others, thanks for the collaboration and the hilarious times. I'll never forget Scott's quaternions, Samet's olympiad problems, or Megat's power tower magic (which still blows my mind). I have yet to experience teamwork and mutual exploration like we had in the undergrad lounge. Math and physics are fun! I hope the lounge is still happening, that its residents by some miracle still have 24 hour access into the building, that some barefoot Megat-like character is still gaming in the corner, that they have some Andrew Rice imitator for comic relief, and that an esteemed graduate student like Payne still finds time to stoop to the lounge-residents' level. May the lounge live on forever!

Now we come to Marwan Hassan. About five years ago now I wandered into your office saying I liked rivers and wanted to describe sediment transport with statistical physics. I realize now when I said ``statistical physics" your eyes lit up with visions of Einstein, Nakagawa, Yano, and so many other names I now know well. I was lucky to use the right trigger words.
Marwan, thanks for taking the risk on me, for all of the guidance, for lending me hundreds of books and papers, for giving me freedom to explore ideas, and most of all for showing me that it's ok (and implicitly expected) to be human in science.

@@

Thank you Rui Ferreira and Brett Eaton!
Rui, you welcomed me to Portugal, offered me guidance in the laboratory, and educated me on turbulence and hydraulics. Your kindness, physics knowledge, and wide-ranging interests beyond physical science are inspiring. Thanks for the thoughtful comments and advice on my PhD. I'm not forgetting what I owe you! I will always be impressed at your concise way of speaking and unbounded kindness.
Brett, you were the first person to welcome me to UBC, and I'll always appreciate your effort to teach me how to write, interpret flume experiments, and understand simplified modelling in light of real world complexity. Although I'm not sure if I understood your lessons as well as you would have liked, I still have a list of your favorite books on my computer, and they're a first task for as soon as I'm done. Maybe they'll bring me closer to your way of thinking!

I'd be amiss not to mention David Furbish. Before I started my PhD, when I was still exclusively a physics kid, I attended Furbish's colloquium in Geography at Marwan's invitation. At the end I asked a vague question about whether Langevin equations might be useful for describing sediment transport, and the answer was ``absolutely yes". Now I have a thesis on it. David later recruited me into a lovely series of meetings where we discussed gases with adhesion, moon craters, entropy, soft matter, martian hillslopes, the social impacts of disinformation, and a whole range of questions in natural philosophy from stochastic landscape evolution, to probability as a physical quantity like energy and momentum and fringe ideas in cosmology. These are my people! David, Rachel, Sarah W, Sarah Z, Nakul, Tyler, Erika, and Shawn, thanks sincerely for letting me board the spaceship.

@@

5
\endinput

>>>>

Most of all, to mom Calisa Pierce, dad Jim Pierce, my sisters Kelsey and Kim, and my grandmother Wilma Steele, you've formed all of the best parts of my personality and guided me toward this point, and it's impossible to explain how thankful I am for that. Regarding family I will also always be thankful for my great aunt Alice who graciously paid for my undergraduate education.

@@

Thanks also to my professors at Southern WV Community College who fostered my way of thinking and provided many of the tools I used for the research in this thesis.
At SWVCC, I'd like to thank especially Mindy Saunders, Tex Wood, Anne Klein, Charles Wood, Don Saunders, and George Trimble. Mindy built my math and physics foundation with her incredibly effective teaching, Anne showed me there's invisible order in everything, and Tex demonstrated that art will always lead science. SWVCC was a personal transformation and a great privilege.

@@

To my extended cohort at WVU physics, Payne, Megat, Collins, Scott, Dustin, Rice, Wolfe, Samet, Craig, April, Evan, Larry, Robert, Stephen, Rick, Matt, Caleb, John, and all of you others, thanks for the collaboration and the hilarious times. I'll never forget Scott's quaternions, Samet's olympiad problems, or Megat's power tower magic (which still blows my mind). I hope the lounge is still happening, that its residents by some miracle still have 24 hour access into the building, that some barefoot Megat-like character is still gaming in the corner, that they have some Andrew Rice imitator for comic relief, and that an esteemed graduate student like Payne still finds time to stoop to the lounge-residents' level. May the lounge live on forever!

Now we come to Marwan Hassan. Some time ago now I wandered into your office saying I liked rivers and wanted to describe sediment transport with statistical physics. I realize now when I said ``statistical physics" your eyes lit up with visions of Einstein, Nakagawa, Yano, and so many other names I now know well. I was lucky to use the right trigger words.
Marwan, thanks for taking the risk on me, for all of the guidance, for lending me hundreds of books and papers, for giving me freedom to explore ideas, and most of all for showing me that it's ok (and implicitly expected) to be human in science.

@@

Thank you Rui Ferreira and Brett Eaton!
Rui, you welcomed me to Portugal, offered me guidance in the laboratory, and educated me on turbulence and hydraulics. Your kindness, physics knowledge, and wide-ranging interests beyond physical science are inspiring. Thanks for the thoughtful comments and advice on my PhD. I'm not forgetting what I owe you! I will always be impressed at your concise way of speaking and unbounded kindness.
Brett, you were the first person to welcome me to UBC, and I'll always appreciate your effort to teach me how to write, interpret flume experiments, and understand simplified modelling in light of real world complexity. Although I'm not sure if I understood your lessons as well as you would have liked, I still have a list of your favorite books on my computer, and they're a first task for as soon as I'm done. Maybe they'll bring me closer to your way of thinking.

I'd be amiss not to mention David Furbish. Before I started my PhD, when I was still exclusively a physics kid, I attended Furbish's colloquium in Geography at Marwan's invitation. At the end I asked a vague question about whether Langevin equations might be useful for describing sediment transport, and the answer was ``absolutely yes". Now I have a thesis on it. David later recruited me into a lovely series of meetings where we discussed gases with adhesion, moon craters, entropy, soft matter, martian hillslopes, the social impacts of disinformation, and a whole range of questions in natural philosophy. David, Rachel, Sarah W, Sarah Z, Nakul, Tyler, Erika, and Shawn, thanks sincerely for letting me board the spaceship.

@@

\endinput

### 2021-07-07 app_ch2.tex
this is a nightmare

\chapter{Ch. 2 calculations}

\section{Derivation of the master equation}
\section{Solution for the position probability distribution}
\section{Moments of position}
\section{Calculation of the flux rate constant}

>>>>

\chapter{Calculations involved in dynamical sediment flux model}
\label{sec:appendixA}
\section{Derivation of the master equation for particle position}
\label{sec:appAmaster}
Here I derive the master equation \ref{eq:flippymaster} for the probability distribution of particle position from the Langevin equation \ref{eq:flippylangevin}. This is closely based off of the approach in \citet{Balakrishnan1993}.

The master equation satisfies $P(x,t) = \bra\bra \delta(x-x(t))\ket_\xi \ket_\eta$, where the averages are over all realizations of the two independent noises. It is most convenient to take these averages after taking Fourier transforms and manipulating time derivatives of the resulting equations.

Integrating the Langevin equation \ref{eq:flippylangevin}, substituting the solution in the above definition, and taking the Fourier transform in space gives
\be \tilde{P}(g,t) = \Big\bra  \Big\bra \exp \Big[- i g \int_0^t du [V+\sqrt{2D}\xi(u)]\eta(u) \Big]\Big\ket_\eta \Big\ket_\xi\ee
This can be rearranged as
\be \tilde{P}(g,t) = \Big\bra \exp\Big[i g V \int_0^t du \eta(u)\Big] \Big\bra \exp\Big[ i g \sqrt{2D}\int_0^t du \xi(u) \eta(u)]\Big]\Big\ket_\xi \Big\ket_\eta .\ee

Using the classic identity for an average over white noise of an exponential \citep{Balakrishnan1993,VanKampen2007} gives, after recognizing that the dichotomous noise satisfies $\eta^2 = \eta$ since $\eta$ is either $0$ or $1$,
\be \tilde{P}(g,t) = \Big\bra \exp\Big[(igV-g^2D) \int_0^t du \eta(u) \Big]\Big\ket_\eta. \label{eq:logi}\ee

Now I will take time derivatives in order to apply the so-called Furutsu-Novikov formula to evaluate the ensemble average over dichotomous noise \citep{Shapiro1978}. Applied to the dichotomous noise in figure \ref{fig:lislefig}, the Furutsu-Novikov formula is, for an arbitrary functional $F$ of the noise,
\be \pt \bra \eta(t)F[\eta(t)] \ket = \bra \eta \pt F \ket + k \big[\bra \eta \ket \bra F \ket -\bra \eta F \ket \big]  \label{eq:furuti}.\ee

Making the shorthands $G = igV-g^2D$ and $F=\exp G\int_0^t \eta(u)du$ and taking as time derivative of eq. \ref{eq:logi} gives
\be G^{-1}\pt \tilde{P}(g,t) = \bra \eta F\ket. \label{eq:furut}\ee
Taking a second time derivative opens up the possibility of using the Furutsu-Novikov formula \ref{eq:furuti}: 
\be G^{-1}\pt \tilde{P}(g,t) = \pt \bra \eta F\ket = G\bra \eta F \ket + k\big[ \bra \eta \ket \tilde{P}(g,t) - G^{-1} \pt \tilde{P}(g,t) \big]. \ee
Applying eq. \ref{eq:furut} finally gives
\be \pt^2 \tilde{P}(g,t)  = (igV-g^2D-k)\pt  \tilde{P} + k_E (igV-g^2D) \tilde{P},\ee
and inverse Fourier transforming provides the master equation \ref{eq:flippymaster}.

\section{Solution for the position probability distribution}
\label{sec:fluccymastersol}
The position probability distribution can be obtained from eq. \ref{flippymaster} for the initial conditions $P(x,0) = \delta(x)$ and $ \pt P(x,0) = \frac{k_E}{k}\big[D\delta''(x)-V \delta'(x) \big]$ using Fourier transforms in space and Laplace transforms in time. Taking these transforms provides
\be \bar{\tilde{P}}(g,s) = \frac{s + k +\varphi  Dg^2 - i g V \varphi}{s(s+k)+(Dg^2-igV)(s+k_E)}, \ee
where $\varphi = k_D/k$ is the probability a particle starts at rest.

The numerator terms encode the initial conditions. The denominator terms are where the real structure of the solution is contained.
The numerator terms involving $g$ are easily expressed as first and second derivatives with respect to $x$. The problem, then is to calculate the inverse Fourier integral of the denominator.
This can be conducted by finding the zeros of the denominator, expanding in partial fractions, then applying the contour integral \citep[e.g.][]{Arfken1985}
\be \int_{-\infty}^\infty \frac{1}{2\pi i} \frac{e^{-i g x}}{g + i c} dg = \theta(x)e^{-c x}.\ee

Rearranging the governing equation gives
\be \tilde{\hat{P}}(g,s) = \frac{\varphi D g^2 - ig V\varphi  + s + k}{D(s+k_E)}\frac{1}{g^2-i\frac{V}{D}g + \frac{s(s+k)}{D(s+k_E)}}\ee
The roots of the denominator are at
\be g_\pm = \frac{iV}{2D}\Big[ 1  \pm R \Big],\ee
where
\be R = \sqrt{1 + \frac{4D}{V^2}\frac{s(s+k)}{s+k_E}}\ee
The partial fractions expansion is therefore
\be \frac{1}{g^2-i\frac{V}{D}g + \frac{s(s+k)}{D(s+k_E)}} = \frac{D}{i V R}\Big[ \frac{1}{g-g_+} - \frac{1}{g-g_-}\Big]\ee
Giving 
\be \tilde{P}(x,s) = \frac{-\varphi D\px^2 + V\varphi\px + s + k}{VR(s+k_E)}\Bigg[ \int \frac{dg}{2\pi i}\frac{e^{-igx}}{g-g_+} - \int \frac{dg}{2\pi i}\frac{e^{-igx}}{g-g_-}\Bigg]\ee
or eventually
\be \tilde{P}(x,s) = \frac{-\varphi D\px^2 + V\varphi \px + s + k}{VR(s+k_E)}\Bigg[ \theta(x) \exp\Big(\frac{Vx}{2D}(1-R)\Big) + \theta(-x)\exp\Big( \frac{Vx}{2D}(1+R)\Big) \Bigg].\ee
This simplifies slightly to
\be \tilde{P}(x,s) = \frac{-\varphi D\px^2 + V\varphi \px + s + k}{VR(s+k_E)}\exp\Big[\frac{Vx}{2D} - \frac{V|x|}{2D}R \Big], \label{eq:laplaceapp}\ee
where as a reminder, $\varphi = k_D/k$.

Taking the inverse Laplace transform symbolically provides the desired probability density function
\be P(x,t) = \mathcal{L}^{-1}\Bigg\{ \frac{1}{V} \big[-\varphi D\px^2 + V\varphi \px + k + s \big] \exp\Big(\frac{Vx}{2D}\Big) \frac{\exp\Big(-\frac{V|x|R}{2D} \Big)}{(s+k_E)R} \Bigg\}(t). \ee
Using the shift property of Laplace transforms \citep[e.g.][]{Arfken1985}, this becomes
\be P(x,t) = \frac{1}{V}e^{-k_E t} \mathcal{L}^{-1}\Bigg\{ \big[-\varphi D\px^2 + V \varphi \px + k_D + s\big] \exp\Big(\frac{Vx}{2D}\Big) \frac{\exp\Big(\frac{-V|x|R_\ast}{2D}\Big)}{s R_\star} \Bigg\},\ee
where 
\be R_\ast = \sqrt{1 + \frac{4D(k_D-k_E)}{V^2}+ \frac{4D}{V^2}\Big(s-\frac{k_Ek_D}{s}\Big)}. \ee
Using the property that $\mathcal{L}^{-1}\big\{s \tilde{f}\big\} = (\delta(t) + \pt)f(t)$ \citep{Arfken1985} gives
\be P(x,t) = \frac{1}{V} e^{-k_E t} \big[ -\varphi D\px^2 + V\varphi \px + k_D + \delta(t) + \pt \big] \exp\Big[\frac{Vx}{2D}\Big] \mathcal{L}^{-1}\Bigg\{ \frac{-\exp\Big[\frac{-V|x|R_\ast}{2D}\Big]}{sR_\ast} \Bigg\}(t) \ee
Finally, using the identity
\be \mathcal{L}^{-1} \Big\{ \frac{1}{s} \tilde{g}(s-a/s)\Big\} = \int_0^t \mathcal{I}_0\Big(2\sqrt{au(t-u)}\Big) g(u) du
\ee
from \citet{Bateman1953}, pg. 133 (which can be derived by either u-substitution or series expansion in powers of $a$) provides:
\begin{multline}  P(x,t) = \frac{1}{V} e^{-k_E t}  \big[ -\varphi D\px^2 + V\varphi \px + k_D + \delta(t) + \pt \big] \exp\Big[\frac{Vx}{2D}\Big]\\ \times \int_0^t du \mathcal{I}_0\Big( 2 \sqrt{k_Ek_D u(t-u)}\Big) \\ \times \mathcal{L}^{-1}\Bigg\{ \frac{\exp\Big[ \frac{-V|x|}{2D}\sqrt{a+b s}\Big]}{\sqrt{a + b s}}\Bigg\}(u)  \label{eq:magic}\end{multline}
where $a = 1 + 4D(k_E-k_E)/(V^2)$ and $b = 4D/V^2$ are shorthands.

Using the trick of introducing a parameter and integrating over it to remove the square root from the denominator, the remaining transform can be evaluated from standard tables \citep{Prudnikov1992a}, eventually giving (after integration over and setting to 1 of the fake parameter),
\begin{multline} P(x,t) = \big[-\varphi D\px^2 + V\varphi \px + k + \delta(t)+ \pt \big]\int_0^t \mathcal{I}_0\Big( 2 \sqrt{k_Ek_D u(t-u)}\Big) e^{-k_E(t-u)} \\ \times \sqrt{ \frac{1}{4\pi D u}} \exp\Big[-k_D u - \frac{(x-Vu)^2}{4Du}\Big] du, \end{multline}
as reported in equation \ref{eq:monster}.

\section{Calculation of the scale-dependent rate function $\Lambda(T)$}
\label{sec:fluxconstant}
The central object required to calculate the sediment flux probability distribution in equation \ref{eq:flippyflux} is 
\be \Lambda(T) = \rho \int_0^\infty dx_i \int_0^\infty dx P(x+x_i,T).\ee
This represents the rate at which particles starts at $x_i$, anywhere to the left of $x=0$, and manage to cross $x=0$ by time $T$.

Again it is easiest to conduct the necessary calculus after Laplace transforming. This gives 
\be \tilde{\Lambda}(s) = \rho \int_0^\infty dx_i \int_0^\infty dx \tilde{P}(x+x_i,s).\ee
The Laplace transform of the rate function therefore follows after integrating equation \ref{eq:laplaceapp} twice.

Noting that $x+x_i$ is always positive,
The first integration gives
\begin{multline} \tilde{\Lambda}(s) = \rho \int_0^\infty dx_i \exp\Big[\frac{V(1-R)x_i}{2D}\Big]\\ 
	\times\Big(\varphi \frac{1-R}{2R(s+k_E)} - \frac{\varphi}{R(s+k_E)} - \frac{2D(s+k)}{V^2R(1-R)(s+k_E)}\Big), \end{multline}
and the second integration provides
\begin{multline}  \tilde{\Lambda}(s) = -\frac{\rho \varphi D}{V R(s+k_E)}  + \frac{2\rho D\varphi}{VR(1-R)(s+k_E)} \\ + \frac{4\rho D^2(s+k)}{V^3R(1-R)^2(s+k_E)}. 
	\label{eq:laplacefluxrate}\end{multline}

Taking the inverse transform, converting the $s$ factor in the numerator of the last term to $\pt + \delta(t)$, and using the shift property gives \citep[e.g.][]{Arfken1985}
\be \Lambda(t) = \rho \mathcal{L}^{-1}\Bigg\{ - \frac{\varphi D}{VR_\star s} + \frac{2D\varphi}{VR_\ast(1-R_\star)s} + \frac{4D^2\big( \cev{\pt} + k\big)}{V^3 R_\star(1-R_\star)^2s}\Bigg\}.\ee
Here the notation $\cev{\pt}$ means the derivative acts from the left on all terms multiplying it (as in $f(t)\cev{\pt} g(t) = \pt [f(t)g(t)]$).

After some work, using eq. \ref{eq:magic} and tabulated Laplace transform pairs \citep[e.g.][]{Arfken1985,Prudnikov1992a} to perform the inverse Laplace transforms, the net result for the rate constant is
\begin{multline} 
	\Lambda(t) = \rho \int_0^t \mathcal{I}_0\Big(2\sqrt{k_Ek_Du(t-u)}\Big)e^{-k_E(t-u)-k_D u} \\
	\times \Bigg[\sqrt{\frac{D}{\pi u}}\Big([\cev{\partial_t} + k]u-\frac{k_D}{2 k}\Big)e^{-V^2 u/4D}  \\ + \frac{V}{2}\Big([\cev{\partial_t} + k]u -\frac{k_D}{k}\Big) \erfc\Bigg(-\sqrt{\frac{V^2 u}{4D}}\Bigg)\Bigg] du,
\end{multline}
as reported in equation \ref{eq:fluxy}.

\label{sec:fluxlimits}
Equation \ref{eq:fluxy} is difficult to work with. To explore the behavior at extreme values of the observation time $T$, one can apply Tauberian theorems \citep[e.g.][]{Weiss1994} to invert the Laplace-transformed rate function of equation \ref{eq:laplacefluxrate} at the opposite extreme of $s$.

For example, at short times, expanding equation \ref{eq:laplacefluxrate} as $s\rightarrow \infty$ gives
\be \tilde{\Lambda}(s) = \frac{\rho k_E V}{2 k s^2} + \frac{\rho k_E}{k}\sqrt{\frac{D}{4 s^3}} \ee
which inverts to
\be \Lambda(t) \sim \frac{\rho k_E V T}{2 k} + \frac{\rho k_E}{k}\sqrt{\frac{D T}{\pi}},\ee
giving the small $T$ behavior.
This has two scaling limits within it. Provided that $T \ll 4D/V^2/\pi < 2 D/V^2$, the scaling goes as $\Lambda(T) \sim T^{-1/2}$. But if $T\gg 4D/V^2/\pi$, it goes as $\Lambda(t) \sim T$.

For large times, taking $s\rightarrow 0 $ gives
\be \tilde{\Lambda}(s) = \frac{\rho k_E V }{ks^2}, \ee
and this inverts to $\Lambda(T) = \rho k_E V T/k$ as the long time solution, providing a mean flux equivalent to the Einstein model (sec. \ref{sec:lisle}).
These limits are summarized in equation \ref{eq:fluccolimitcases}.

### 2021-07-07 ch1intro.tex
this is a nightmare

\subsection{Birth death models for bedload flucuations}

The prevalence of large bedload fluctuations motivated \citet{Ancey2006,Ancey2008} to revisit Einstein's assumptions to develop a model of the bedload flux as a random process.
They derived the probability distribution of the flux by counting the number of moving particles in a control volume, considering that this number changes through time as a result of particles migrating into the volume from upstream, entraining, depositing, and migrating out of the volume to downstream.
To obtain realistically-wide fluctuations in particle activity, they introduced a positive feedback called ``collective entrainment", whereby the entrainment rate of grains increases in proportion to the number of moving grains in the volume \citep{Ancey2008,Heyman2013}.

>>>>

\subsection{Birth death models for bedload flucuations}
\label{sec:birthdeath}
The prevalence of large bedload fluctuations motivated \citet{Ancey2006,Ancey2008} to revisit Einstein's assumptions to develop a model of the bedload flux as a random process.
They derived the probability distribution of the flux by counting the number of moving particles in a control volume, considering that this number changes through time as a result of particles migrating into the volume from upstream, entraining, depositing, and migrating out of the volume to downstream.
To obtain realistically-wide fluctuations in particle activity, they introduced a positive feedback called ``collective entrainment", whereby the entrainment rate of grains increases in proportion to the number of moving grains in the volume \citep{Ancey2008,Heyman2013}.
